$Z^\prime$ mass limits and the naturalness of supersymmetry

### 2015-05-17 paper/e6ssm-tuning-paper.tex
removed spaces bewtween words and footnotes as this looked weird.

However if the extra $U(1)$ gauge symmetries appears from the breakdown of
$E_6$, then $\sin\chi$ should be zero at the GUT scale.  Nonetheless even
if this term is zero at the outset it will still be radiatively generated
if the trace of their charges, $\sum_i Q_i^Y Q^\prime_i$, is non-zero.  In
the cases studied here the trace of the charges over states in the complete
$27$-plets vanish, but to be consistent with single step gauge coupling
unification we also included $\hat{H}^\prime$ and $\hat{\overline{H^\prime}}$
which lead to a non-zero value for $\sum_i Q_i^Y Q^\prime_i $.  The value
induced by this at the EW scale though is rather small as can be seen
\footnote{The specific incomplete multiplets we consider here correspond
to the third of the four possible embeddings they refer to.} in Fig.~3 of
\cite{Rizzo:1998ut}, and this was also checked with two-loop RGEs in the
E$_6$SSM \cite{King:2005jy,Athron:2009bs}.  For this reason and due to the
huge expansion in the number of terms in the two-loop RGEs when one allows
for gauge kinetic mixing we will neglect this in our analysis here and
throughout this paper.

@@

To illustrate this, in the left panel of Fig.~\ref{Fig:mssmstoptuning}
we show the variation in fine tuning for $M_X = 20$ TeV, $50$ TeV, $100$
TeV and $10^{16}$ GeV when we scan over the stop masses and mixing,
with $500 \textrm{ GeV} \leq m_{Q_3} , m_{u_3} \leq 10 \textrm{ TeV}$ and
$-3810\textrm{ GeV} \leq A_t \leq -20 \textrm{ GeV}$.  The remaining
parameters we fix such that at $M_{\rm SUSY}$ they have the values
$\mu = -97.5$ GeV, $B = -84.8$ GeV, $M_1 = 92.1$ GeV, $M_2 = 95.9$ GeV,
$M_3 = 352$ GeV, $A_b = -117.9$ GeV, $A_\tau = -7.8$ GeV, $m_{L_i} = 400$ GeV,
$m_{e_i} = 204$ GeV, $m_{Q_{1,2}} = 438$ GeV, $m_{u_{1,2}} = 436$ GeV and
$m_{d_i} = 438$ GeV ($i = 1 , 2 , 3$).  Here we denote by $M_1$, $M_2$ and
$M_3$ the soft gaugino masses for $U(1)$, $SU(2)$ and $SU(3)$, while $A_b$ and
$A_\tau$ are soft trilinear couplings and the $m_\phi$ are soft scalar
masses for the indicated fields.  The soft bilinear coupling $B$ is defined
such that at tree level the mass of the CP-odd MSSM Higgs boson reads
$m_A^2 = 2 B \mu / \sin 2\beta$.  All off-diagonal couplings and scalar masses
are set to zero, as are the first and second generation Yukawa couplings
and soft trilinears.  Although we should stress that making
this choice will lead to a spectrum which is in conflict with the LHC
limits, doing so ensures that fine tuning due to the other parameters
is small, so that we avoid washing out the fine tuning impact of the
stops when the tuning is small\footnote{For models in which the
spectrum is heavier, when the stop masses are small the fine tuning
reaches a lower bound imposed by other heavier parameters.} as can
be the case when the stop masses are less than 1 TeV.  Note that the
Higgs mass is also allowed to vary in this scan, as shown in the right
panel of Fig.~\ref{Fig:mssmstoptuning}.  This illustrates the tuning
problem which people have been worrying about since the discovery of
the 125 GeV Higgs boson as we see that raising the stop masses is also
pushing up the Higgs mass, meaning that heavier Higgs masses require
more fine tuning.  However for a low value of the UV scale this tuning
is not so severe unless the stops are very heavy, and a 125 GeV Higgs
can be obtained without much tuning in this unrealistic case where we
have minimized other sources of tuning.  On the other hand the tuning
becomes more severe as we increase the cutoff such that for $M_X =
10^{16}$ GeV a lightest stop masses of $1-3$ TeV can result in a fine
tuning of $\approx 100 - 1000$ and the minimum tuning we find
\footnote{Note that in the calculation of the Higgs mass there is a
significant theoretical error, even with leading two-loop corrections,
which should be considered when thinking about what the results imply for
the minimum fine tuning in the model consistent with the recent discovery
of a 125 GeV Higgs.} for a $125$ GeV Higgs is $\approx 200$, as shown in
Fig.~\ref{Fig:mssmstoptuning}.

@@

There is another limit which plays a similar role.  Chargino limits
directly constrain the $\mu$ parameter (or effective $\mu$ parameter
in these $U(1)$ extensions).  The LEP bound \cite{Kraan:2005vy} on
chargino masses, excluding $m_{\tilde{\chi}^\pm_1} \lesssim 104$ GeV, implies
that $|\mu|$ should only be greater than $\sim 100$ GeV, which is not
substantially larger than $M_Z$.  Consequently the bound from LEP is not
high enough to have an impact on the fine tuning obtained in the models and
parameter space regions that we have studied, as we have checked explicitly
\footnote{It is for this reason that we did not require the LEP bounds be
satisfied for the points shown in Figure \ref{Fig:e6ssmvsmssm}
and in Figure \ref{Fig:othere6modelsvsmssm} below.}.  Significantly larger
lower bounds on the $\mu$ parameter, and therefore on the fine tuning, may
arise from chargino limits coming from LHC searches.  However the chargino
limits from the LHC depend on whether there are light sleptons or sneutrinos
and the mass difference between the lightest chargino and lightest neutralino.
Current limits placed by CMS and ATLAS extend up to
$m_{\tilde{\chi}^\pm_1} \approx 700 - 740$ GeV if there are light
sleptons \cite{Khachatryan:2014qwa, Aad:2014nua} with much weaker
bounds if there are no light sleptons or sneutrinos\footnote{Useful
summary plots of these limit may be found on the public pages of
ATLAS, \url{https://atlas.web.cern.ch/Atlas/GROUPS/PHYSICS/CombinedSummaryPlots/SUSY/ATLAS_SUSY_EWSummary/ATLAS_SUSY_EWSummary.png}
and CMS \url{http://cms.web.cern.ch/sites/cms.web.cern.ch/files/styles/large/public/field/image/Image_03_exclusion_Combined.png?itok=8FMBpu_1}.}.

@@

Interestingly other sensitivities are suppressed by this effect since
at the same time larger $M_3$ at higher scales reduces the soft squark
masses at $M_X$.  Therefore the stop mass contributions are
ameliorated, compared to the MSSM, both by allowing lighter stops at
$M_{SUSY}$ and by the modified RGE running.  Nonetheless the stops
still do lead to $\Delta_{m_{H_u}^2}$ increasing with the cutoff
through the usual mechanism \footnote{Wherein $m_{H_u}^2(M_{SUSY})$
receives a positive contribution to it's mass from $m_{H_u}^2(M_X)$
and a negative contribution from $m_{Q_3}^2(M_X)$ and
$m_{u_3}^2(M_X)$, allowing heavy stop masses to cause fine
tuning.  In this case $m_{H_u}^2(M_{SUSY})$ is held fixed so as the
cutoff increases the values of these soft masses at $M_X$ will be
larger and there will be a bigger cancellation between them,
increasing the sensitivity of $M_Z$ to both $m_{H_u}^2$ and the soft
scalar masses for the stops.}.

@@

When this trace is non-zero it will also induce gauge kinetic mixing
to be generated during RGE evolution and this is the case in the
models we consider here \footnote{In $E_6$ inspired models with only
complete $27$-plet matter multiplets this trace would vanish.
However since we assume some incomplete multiplets so that our
models are consistent with gauge coupling unification this trace
doesn't vanish.}.  However when these models are evolved down from
the GUT scale the radiatively generated gauge kinetic mixing gives an
off diagonal gauge coupling, $g_{11}$, of just $\approx 0.02$ at the
electroweak scale \cite{King:2005jy} and so it does not play a large
role.  At the same time if gauge kinetic mixing is included the RGE
expressions become very large and unmanageable, so we neglected the
gauge kinetic mixing by setting the SARAH flag \texttt{NoU1Mixing} to true.

@@

with \footnote{The first of these is the trace which is assumed to vanish in

>>>>

However if the extra $U(1)$ gauge symmetries appears from the
breakdown of $E_6$, then $\sin\chi$ should be zero at the GUT scale.
Nonetheless even if this term is zero at the outset it will still be
radiatively generated if the trace of their charges, $\sum_i Q_i^Y
Q^\prime_i$, is non-zero.  In the cases studied here the trace of the
charges over states in the complete $27$-plets vanish, but to be
consistent with single step gauge coupling unification we also
included $\hat{H}^\prime$ and $\hat{\overline{H^\prime}}$ which lead
to a non-zero value for $\sum_i Q_i^Y Q^\prime_i $.  The value induced
by this at the EW scale though is rather small as can be
seen\footnote{The specific incomplete multiplets we consider here
  correspond to the third of the four possible embeddings they refer
  to.} in Fig.~3 of \cite{Rizzo:1998ut}, and this was also checked
with two-loop RGEs in the E$_6$SSM \cite{King:2005jy,Athron:2009bs}.
For this reason and due to the huge expansion in the number of terms
in the two-loop RGEs when one allows for gauge kinetic mixing we will
neglect this in our analysis here and throughout this paper.

@@

To illustrate this, in the left panel of Fig.~\ref{Fig:mssmstoptuning}
we show the variation in fine tuning for $M_X = 20$ TeV, $50$ TeV,
$100$ TeV and $10^{16}$ GeV when we scan over the stop masses and
mixing, with $500 \textrm{ GeV} \leq m_{Q_3} , m_{u_3} \leq 10
\textrm{ TeV}$ and $-3810\textrm{ GeV} \leq A_t \leq -20 \textrm{
  GeV}$.  The remaining parameters we fix such that at $M_{\rm SUSY}$
they have the values $\mu = -97.5$ GeV, $B = -84.8$ GeV, $M_1 = 92.1$
GeV, $M_2 = 95.9$ GeV, $M_3 = 352$ GeV, $A_b = -117.9$ GeV, $A_\tau =
-7.8$ GeV, $m_{L_i} = 400$ GeV, $m_{e_i} = 204$ GeV, $m_{Q_{1,2}} =
438$ GeV, $m_{u_{1,2}} = 436$ GeV and $m_{d_i} = 438$ GeV ($i = 1 , 2
, 3$).  Here we denote by $M_1$, $M_2$ and $M_3$ the soft gaugino
masses for $U(1)$, $SU(2)$ and $SU(3)$, while $A_b$ and $A_\tau$ are
soft trilinear couplings and the $m_\phi$ are soft scalar masses for
the indicated fields.  The soft bilinear coupling $B$ is defined such
that at tree level the mass of the CP-odd MSSM Higgs boson reads
$m_A^2 = 2 B \mu / \sin 2\beta$.  All off-diagonal couplings and
scalar masses are set to zero, as are the first and second generation
Yukawa couplings and soft trilinears.  Although we should stress that
making this choice will lead to a spectrum which is in conflict with
the LHC limits, doing so ensures that fine tuning due to the other
parameters is small, so that we avoid washing out the fine tuning
impact of the stops when the tuning is small\footnote{For models in
  which the spectrum is heavier, when the stop masses are small the
  fine tuning reaches a lower bound imposed by other heavier
  parameters.} as can be the case when the stop masses are less than 1
TeV.  Note that the Higgs mass is also allowed to vary in this scan,
as shown in the right panel of Fig.~\ref{Fig:mssmstoptuning}.  This
illustrates the tuning problem which people have been worrying about
since the discovery of the 125 GeV Higgs boson as we see that raising
the stop masses is also pushing up the Higgs mass, meaning that
heavier Higgs masses require more fine tuning.  However for a low
value of the UV scale this tuning is not so severe unless the stops
are very heavy, and a 125 GeV Higgs can be obtained without much
tuning in this unrealistic case where we have minimized other sources
of tuning.  On the other hand the tuning becomes more severe as we
increase the cutoff such that for $M_X = 10^{16}$ GeV a lightest stop
masses of $1-3$ TeV can result in a fine tuning of $\approx 100 -
1000$ and the minimum tuning we find\footnote{Note that in the
  calculation of the Higgs mass there is a significant theoretical
  error, even with leading two-loop corrections, which should be
  considered when thinking about what the results imply for the
  minimum fine tuning in the model consistent with the recent
  discovery of a 125 GeV Higgs.} for a $125$ GeV Higgs is $\approx
200$, as shown in Fig.~\ref{Fig:mssmstoptuning}.

@@

There is another limit which plays a similar role.  Chargino limits
directly constrain the $\mu$ parameter (or effective $\mu$ parameter
in these $U(1)$ extensions).  The LEP bound \cite{Kraan:2005vy} on
chargino masses, excluding $m_{\tilde{\chi}^\pm_1} \lesssim 104$ GeV, implies
that $|\mu|$ should only be greater than $\sim 100$ GeV, which is not
substantially larger than $M_Z$.  Consequently the bound from LEP is not
high enough to have an impact on the fine tuning obtained in the models and
parameter space regions that we have studied, as we have checked explicitly\footnote{It is for this reason that we did not require the LEP bounds be
satisfied for the points shown in Figure \ref{Fig:e6ssmvsmssm}
and in Figure \ref{Fig:othere6modelsvsmssm} below.}.  Significantly larger
lower bounds on the $\mu$ parameter, and therefore on the fine tuning, may
arise from chargino limits coming from LHC searches.  However the chargino
limits from the LHC depend on whether there are light sleptons or sneutrinos
and the mass difference between the lightest chargino and lightest neutralino.
Current limits placed by CMS and ATLAS extend up to
$m_{\tilde{\chi}^\pm_1} \approx 700 - 740$ GeV if there are light
sleptons \cite{Khachatryan:2014qwa, Aad:2014nua} with much weaker
bounds if there are no light sleptons or sneutrinos\footnote{Useful
summary plots of these limit may be found on the public pages of
ATLAS, \url{https://atlas.web.cern.ch/Atlas/GROUPS/PHYSICS/CombinedSummaryPlots/SUSY/ATLAS_SUSY_EWSummary/ATLAS_SUSY_EWSummary.png}
and CMS \url{http://cms.web.cern.ch/sites/cms.web.cern.ch/files/styles/large/public/field/image/Image_03_exclusion_Combined.png?itok=8FMBpu_1}.}.

@@

Interestingly other sensitivities are suppressed by this effect since
at the same time larger $M_3$ at higher scales reduces the soft squark
masses at $M_X$.  Therefore the stop mass contributions are
ameliorated, compared to the MSSM, both by allowing lighter stops at
$M_{SUSY}$ and by the modified RGE running.  Nonetheless the stops
still do lead to $\Delta_{m_{H_u}^2}$ increasing with the cutoff
through the usual mechanism\footnote{Wherein $m_{H_u}^2(M_{SUSY})$
receives a positive contribution to it's mass from $m_{H_u}^2(M_X)$
and a negative contribution from $m_{Q_3}^2(M_X)$ and
$m_{u_3}^2(M_X)$, allowing heavy stop masses to cause fine
tuning.  In this case $m_{H_u}^2(M_{SUSY})$ is held fixed so as the
cutoff increases the values of these soft masses at $M_X$ will be
larger and there will be a bigger cancellation between them,
increasing the sensitivity of $M_Z$ to both $m_{H_u}^2$ and the soft
scalar masses for the stops.}.

@@

When this trace is non-zero it will also induce gauge kinetic mixing
to be generated during RGE evolution and this is the case in the
models we consider here\footnote{In $E_6$ inspired models with only
complete $27$-plet matter multiplets this trace would vanish.
However since we assume some incomplete multiplets so that our
models are consistent with gauge coupling unification this trace
doesn't vanish.}.  However when these models are evolved down from
the GUT scale the radiatively generated gauge kinetic mixing gives an
off diagonal gauge coupling, $g_{11}$, of just $\approx 0.02$ at the
electroweak scale \cite{King:2005jy} and so it does not play a large
role.  At the same time if gauge kinetic mixing is included the RGE
expressions become very large and unmanageable, so we neglected the
gauge kinetic mixing by setting the SARAH flag \texttt{NoU1Mixing} to true.

@@

with\footnote{The first of these is the trace which is assumed to vanish in
